Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2021-07-11 ms/Thesis/thesis.tex
Go over Yoav's comments

Beside kin selection, two more major theroies were suggested to explain to evolution of cooperation.

\textbf{Reciprocity} suggests repeating interactions or individual recognition as key factors for explaining the evolution of cooperation. 
In \textit{direct reciprocity} there are a repeated encounters between the same two individuals. In every encounter, each player has a choice between cooperation and defection. If I cooperate now, you may cooperate later. Hence, it may pay off to cooperate.
This game-theoretic framework is known as the repeated Prisoner's Dilemma. 
Direct reciprocity can only lead to the evolution of cooperation if the cost is smaller than $w$ the probability for another encounter between the same two individuals multiplied by the benefit. 
\begin{equation} \label{reciprocity}
c<bw
\end{equation}
Direct reciprocity assumes that both players are in a position to cooperate. Direct reciprocity can not explain cooperation in asymmetric interactions. In humans, such interactions happen often, for example humans often donate money. 
\textit{Indirect reciprocity} has been suggested to explain this behavior. 
Nowak \cite{nowak2006five} claims that direct reciprocity is like a barter economy based on the immediate exchange of goods, while indirect reciprocity resembles the invention of money. The money that "fuels the engines" of indirect reciprocity is reputation. 
However, Reciprocity assume repeating interactions and therefore, has difficulty explaining evolution of cooperation if the no repeating interactions occurs.

\textbf{Group Selection} theory posits that cooperation is favoured because of the advantage to the whole group, if selection acts at the group level in addition to the individual level. A common model for group selection work as is: the population is divided into groups. In each groups there are cooperators, which help to other group members and defectors which do not help. 
Individuals reproduce proportional to their fitness. Offspring are added to the same group.
If a group reaches a certain size it can split to two groups. A group that grow faster will split more often. Groups of cooperators are growing faster than group of defectors.
Therefore, cooperation can evolve in this model when the ratio between benefit b and cost c is more than one plus the ratio between the maximum group size n and the number of groups m:
\begin{equation} \label{groupselection}
\frac{b}{c}>1+\frac{n}{m}
\end{equation}

All three theories mentioned above assume that cooperation is genetically determined. This raise the question, is it possible that cooperation is determined by environmental or social influences.
Cooperative behavior can be subject to \emph{cultural transmission}, which allows an individual to acquire attitudes or behavioral traits from other individuals in its social group through imitation, learning, or other modes of communication \citep{cavalli1981cultural,richerson2008not}.
Cultural transmission may be  modeled as vertical, horizontal, or oblique:  vertical transmission occurs between parents and offspring, horizontal transmission occurs between individuals from the same generation, and oblique transmission occurs  to offspring from the generation to which their parents belong (i.e. from non-parental adults). 
Evolution under either of these transmission models can be be more rapid than under pure vertical transmission~\citep{cavalli1981cultural,lycett2008questions,ram2018evolution}.

To understand the evolution of cooperation we are going to use \emph{replicator dynamics}. The replicator in replicator dynamics has the ability to make one or more copies of itself.
The replicator can be a gene, a phenotype, a strategy in a game and etc. In evolutionary game theory context replicator is a different strategy in the game. For example, in the case of cooperation whether the individual is a cooperative or a defector.
In replicator dynamics we assume large population of replicators, which interact with respect to their proportion.
Those interactions of different replicator affect the fitness acorrding to some payoff matrix. The payoff matrix depands on the game which is played. 
The most common game to describe cooperation is prisoner's dilemma\citep{nowak1992tit}.
Similar to dominant strategies bringing forth Nash equilibria when games are repeated, strategies in replicator dynamics can become evolutionary stable. Such strategies are called \emph{Evolutionarily Stable strategy (ESS)}.
Such strategies cannot be invaded by any other strategy that is initially rare. Evolutionarily stable strategy maximizes the expected fitness of its chooser and therefore, causing maximum mean fitness.
An interesting and fundamental question is to verify whether natural selection, operating within the framework of known genetical models leads to ESS.
The answer to the question above is that natural selection usually does not lead to an increase in the average mean fitness of the population. Therefore, it is not generally true that natural selection does operate as to maximize the average individual fitness resulting in an ESS.
\citet{sacks1967stable} even showed that natural selection at a diploid locus model when relative fitnesses of the different genotypes are functions of gene frequency leads to stable equilibrium gene frequency that corresponds to minimum mean fitness.
Instead of searching evolutionarily stable strategy, we focus on \emph{Evolutionary Genetic Stblility (EGS)}\citep{lessard1990evolutionary}. EGS is when the frequency of types (genotype, phenotype etc.) stays the same.
Here, we study both local stability and external stability. Equilibrium is consider \emph{Local stability} when a system near the equilibrium approaches it.
\emph{External stability}\citep{Liberman1986modifiers,Altenberg2017} is a local stability of the equilibrium to invasion by the modifier allele m. One of the main questions in the evolution of cooperation is under what conditions such invasions are possible, or in other words under what conditions the system is external unstable.

@@

\citet{handley2020human} showed the importance of culture on the human behavior. They showed that the probablity of indvidual to cooperate with unrelated strangers from a different group in transient interactions corresponds to the degree of cultural similarity between those groups.
Therefore, they have suggested that group-level selection on culturally differentiated populations can explain cooperation between unrelated humans from different groups.

@@

\subsection{Basic Model}
\label{section:basicmodel}
Consider a large population whose members can be one of two phenotypes: $\phi=A$ for cooperators or $\phi=B$ for defectors.
An offspring inherits its phenotype from its parent via vertical transmission with probability $v$ or from a random individual in the parental population via oblique transmission with probability $(1-v)$ (\autoref{fig:illustration}a). 
Following~\citet{ram2018evolution}, given that the parent's phenotype is $\phi$ and assuming uni-parental inheritance~\citep{Zefferman2016}, the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is 
\begin{equation} \label{eq:vertical_oblique_transmission}
P(\phi'=A \mid \phi) = \begin{cases}
v + (1-v)p, & \text{if } \phi=A \\
(1-v)p, & \text{if } \phi=B
\end{cases},
\end{equation}
where $p=P(\phi=A)$ is the frequency of $A$ among all adults in the parental generation.

@@

\subsection{Invasion Model}
In this section, we use the model described in \autoref{section:basicmodel}. 
Here we assume a coexistence of both phenotypes (cooperators and defectors) and that the frequency of each phenotye stays the same each generation.
Henece, we assume that the we start from one of the equilibrium.
In addition, we assume that the equilibria is stable.  Denote the frequency of cooperators in the population in $\hat{p}^*$ and frequency of defectors $1-\hat{p}^*$.

Starting from the assumption described above we want to investigate the evolution of interaction-transmission association $\alpha$.
We assume that the initial population has interaction-transmission association $\alpha_1$. 
We want to answer the question what would happen to cooperators frequency if a new mutation would occur and change the interaction-transmission association to a different value $\alpha_2$ without direct effect on fitness.
We can think of this model as an invasion model where the invadors are the individuals with interaction-transmission association $\alpha_2$. \autoref{fig:invasion_illustration} illustrates the invasion.

%%% Figure: model illustration - invasion model
\begin{figure}[pt]
  \centering
  \includegraphics[width=0.7\textwidth]{../PRSB_figures/fig_2.pdf}
  \caption{\textbf{Invasion model illustration.} Starting from stable equilibrium with coexistence of both phenotypes. 
  A mutation that has effect only the interaction-transmission association tries to invade. }
  \label{fig:invasion_illustration}
\end{figure}

@@

Our deterministic model provides a good approximation to outcomes of simulations of a complex stochastic model with population structure in which individuals can only interact with and transmit to their neighbors.
In these structured populations interaction-transmission association arises due to both social interactions and horizontal cultural transmission being local (\autoref{fig:spatial} and \autoref{fig:spatial_local_selection}).
We did not find any significant difference between local and global selection. %DC: can we disucss why there is not difference?

@@

Another mechanism that was suggested by \citet{traxler2011social} showed that \emph{conditional cooperation} based on norm-dependent relational utilities, i.e. individual will only cooperate if it knows that others will cooperate too, can sustain cooperation in a community – provided that cooperation is already at a high level.
Unlike \emph{conditional cooperation}, interaction-transmission association can sustain cooperation in community even if cooperation is initially rare. 
\citet{morsky2020false} suggested that false beliefs on the frequencies of the cooperator can affect the individual decision whether to cooperate or not. If the individual over estimate the number of cooperators it will be more likely to cooperate. This false belief can help sustain cooperation even if cooperartion is initially rare. 
% DC: Let's discuss this paragraph - is it good?

>>>>

Beside kin selection, two other major theroies were suggested to explain to evolution of cooperation.

\textbf{Reciprocity} suggests repeating interactions or individual recognition as key factors for explaining the evolution of cooperation. 
In \textit{direct reciprocity} there are a repeated encounters between the same two individuals. In every encounter, each player has a choice between cooperation and defection. If I cooperate now, you may cooperate later. Hence, it may pay off to cooperate.
This game-theoretic framework is known as the \emph{repeated Prisoner's Dilemma}. 
Direct reciprocity can only lead to the evolution of cooperation if the cost is smaller than $w$ the probability for another encounter between the same two individuals multiplied by the benefit, 
\begin{equation} \label{reciprocity}
c<b\cdot w \,.
\end{equation}
Direct reciprocity assumes that both players are in a position to cooperate. Direct reciprocity can not explain cooperation in asymmetric interactions\citep{su2021evolution}. In humans, such interactions happen often, for example, humans donate money. 
\textit{Indirect reciprocity} has been suggested to explain this behavior. 
Nowak \cite{nowak2006five} claims that direct reciprocity is like a barter economy based on the immediate exchange of goods, while indirect reciprocity resembles the invention of currercy. The currercy that "fuels the engines" of indirect reciprocity is \emph{reputation}. 
However, reciprocity assumes repeating interactions and therefore, has difficulty explaining evolution of cooperation if no repeating interactions occurs.

\textbf{Group Selection} theory posits that cooperation is favoured because of the advantage to the whole group, if selection acts at the group level in addition to the individual level. A common model for group selection work as is: the population is divided into groups. In each group there are cooperators, which help other group members, and defectors, which do not help. 
Individuals reproduce proportional to their fitness. Offspring are added to the same group.
If a group reaches a certain size it can split to two groups. A group that grows faster will split more often. Groups of cooperators grow faster than groups of defectors.
Therefore, cooperation can evolve in this model when
\begin{equation} \label{groupselection}
\frac{b}{c}>1+\frac{n}{m}\, ,
\end{equation}
where \emph{n} is the maximum group size and \emph{m} is the number of groups.
\\

All three theories mentioned above assume that cooperation is genetically determined. This raise the question, is it possible that cooperation is determined by environmental or social influences.
Cooperative behavior may be subject to \emph{cultural transmission}, which allows an individual to acquire attitudes or behavioral traits from other individuals in its social group through imitation, learning, or other modes of communication \citep{cavalli1981cultural,richerson2008not}.
Cultural transmission may be  modeled as vertical, horizontal, or oblique:  vertical transmission occurs between parents and offspring, horizontal transmission occurs between individuals from the same generation, and oblique transmission occurs  to offspring from the generation to which their parents belong (i.e. from non-parental adults). 
Evolution under either of these transmission models can be be more rapid than under pure vertical transmission~\citep{cavalli1981cultural,lycett2008questions,ram2018evolution}.

@@

\citet{handley2020human} studied the importance of culture on human behavior. They showed that the probablity of indvidual to cooperate with unrelated strangers from a different group in transient interactions corresponds to the degree of cultural similarity between those groups.
Therefore, they have suggested that group-level selection on culturally differentiated populations can explain cooperation between unrelated humans from different groups.

To understand the evolution of cooperation we are going to use \emph{replicator dynamics}. The replicator in replicator dynamics has the ability to make one or more copies of itself.
The replicator can be a gene, a phenotype, a strategy in a game and etc. In evolutionary game theory context replicator is a different strategy in the game. For example, in the case of cooperation whether the individual is a cooperative or a defector.
In replicator dynamics we assume large population of replicators, which interact with respect to their frequency.
Interactions of different replicator affect the fitness acorrding to some payoff matrix. This payoff matrix depands on the game which is played. 
The most common game to describe cooperation is the \emph{Prisoner's Dilemma}\citep{nowak1992tit}.
Similar to dominant strategies bringing forth Nash equilibria when games are repeated, strategies in replicator dynamics can become evolutionary stable. Such strategies are called \emph{evolutionarily stable strategies (ESS)}.
Such strategies cannot be invaded by any other strategy that is initially rare. Evolutionarily stable strategies maximize the expected fitness of its replicators and therefore, maximize the population mean fitness.
A fundamental question is to test if natural selection, operating within the framework of known genetical models, leads to ESS.
The answer to the question above is that natural selection in many cases does not lead to an increase in the population mean fitness. Therefore, it is not generally true that natural selection does operate as to maximize the average individual fitness resulting in an ESS.
\citet{sacks1967stable} even showed that natural selection at a diploid locus model, when relative fitnesses of the different genotypes are functions of gene frequency, leads to stable equilibrium gene frequencies that corresponds to not maximum mean fitness.
Instead of searching for evolutionarily stable strategies, we focus on \emph{evolutionary genetic stblility (EGS)}\citep{lessard1990evolutionary}. EGS occurs when the frequencies of types (genotypes, phenotypes etc.) remains stable.
Here, we study both \emph{local stability} and \emph{external stability}. \emph{Local stability} occurs when a system near the equilibrium wiil approach it.
\emph{External stability} \citep{Liberman1986modifiers,Altenberg2017} is local stability of the equilibrium to invasion by a modifier allele \emph{m}. One of the main questions in the evolution of cooperation is under what conditions such invasions are possible, or in other words under what conditions the system is externally unstable?

@@

\subsection{Basic Model}
\label{section:basicmodel}
Consider a large population whose members can have one of two phenotypes: $\phi=A$ for cooperators or $\phi=B$ for defectors.
An offspring inherits its phenotype from its parent via vertical transmission with probability $v$ or from a random individual in the parental population via oblique transmission with probability $(1-v)$ (\autoref{fig:illustration}a). 
Following~\citet{ram2018evolution}, given that the parent's phenotype is $\phi$ and assuming uni-parental inheritance~\citep{Zefferman2016}, the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is 
\begin{equation} \label{eq:vertical_oblique_transmission}
P(\phi'=A \mid \phi) = \begin{cases}
v + (1-v)p, & \text{if } \phi=A \\
(1-v)p, & \text{if } \phi=B
\end{cases},
\end{equation}
where $p=P(\phi=A)$ is the frequency of $A$ among all adults in the parental generation.

@@

\subsection{Invasion Model}
In this section, we extend the model described in \autoref{section:basicmodel}. 
We assume coexistence of both phenotypes (cooperators and defectors) and that the frequency of each phenotye stays stable over time.
Hence, we assume that the we start from a stable equilibrium. Denote the frequency of cooperators in the population as $\hat{p}^*$ and frequency of defectors as $1-\hat{p}^*$.

We investigate the evolution of interaction-transmission association $\alpha$.
We assume that the initial population has interaction-transmission association $\alpha_1$. 
What would happen to cooperators frequency if a new mutation would occur that changed the interaction-transmission association to a different value $\alpha_2$ without direct effect on fitness?
Here, the invaders are the individuals with interaction-transmission association $\alpha_2$. \autoref{fig:invasion_illustration} illustrates the invasion.

%%% Figure: model illustration - invasion model
\begin{figure}[pt]
  \centering
  \includegraphics[width=0.7\textwidth]{../PRSB_figures/fig_2.pdf}
  \caption{\textbf{Invasion model illustration.} Starting from stable equilibrium with coexistence of both phenotypes. 
  A mutation that affected only the interaction-transmission association $\alpha$ invades the population }
  \label{fig:invasion_illustration}
\end{figure}

@@

Our deterministic model provides a good approximation to outcomes of simulations of a complex stochastic model with population structure in which individuals can only interact with and transmit to their neighbors.
In these structured populations interaction-transmission association arises due to both social interactions and horizontal cultural transmission being local (\autoref{fig:spatial} and \autoref{fig:spatial_local_selection}).
We did not find any significant difference between local and global selection.

Another mechanism that was suggested by \citet{traxler2011social} showed that \emph{conditional cooperation} based on norm-dependent relational utilities, i.e. individual will only cooperate if it knows that others will cooperate too, can sustain cooperation in a community – provided that cooperation is already at a high frequency.
Unlike \emph{conditional cooperation}, interaction-transmission association can sustain cooperation in a community even if cooperation is initially rare. 
\citet{morsky2020false} suggested that false beliefs on the frequencies of the cooperator can affect the individual decision whether to cooperate or not. If the individual over-estimates the number of cooperators it will be more likely to cooperate. This false belief can help sustain cooperation even if cooperartion is initially rare.

@@
